Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-21 paper.tex
added ack
Refactor dtwf code to fit patterns.
Update

Tree sequences encode a complete and concise description of a set of sampled
chromosome's genetic ancestry, and they can also efficiently represent
the genetic variation (extant and ancestral).
As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
sequence encoding includes \emph{site} and \emph{mutation} tables. Each
row in the site table represents some position along the genome, and
each row in the mutation table represents a change of allelic state
at a particular site and node. Since the number of mutations is typically
small, we can usually store the genetic variation for a given variant
site in constant space. This approach of storing the genetic variation
via the genetic ancestry is both very concise and extremely efficient
to process---for example, computing Tajima's $D$ using \tskit\
is several orders of magnitude faster for large sample sizes than
efficient libraries working with the variant matrix
representation~\citep{ralph2020efficiently}.
\jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
here. Same basic idea, but much cruder.}

@@

@@

\subsection*{Finite sites mutations}
\label{sec-mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

\subsection*{Selection}
Another ellaboration of the standard neutral coalescent with recombination
is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
In the context of the coalescent this amounts to creating a structured population
during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
in which lineages may transit between favored and unfavored backgrounds through 
recombination. This allows for many selective sweep scenarios, including recurrent, partial, 
and soft selective sweeps to be simulated efficiently, however in general this
efficiency comes at the cost of flexibility in comparison to forwards in
time simulation.

>>>>

% JK: commenting this out for now as we're covering it in the Mutations
% section.

% Tree sequences encode a complete and concise description of a set of sampled
% chromosome's genetic ancestry, and they can also efficiently represent
% the genetic variation (extant and ancestral).
% As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
% sequence encoding includes \emph{site} and \emph{mutation} tables. Each
% row in the site table represents some position along the genome, and
% each row in the mutation table represents a change of allelic state
% at a particular site and node. Since the number of mutations is typically
% small, we can usually store the genetic variation for a given variant
% site in constant space. This approach of storing the genetic variation
% via the genetic ancestry is both very concise and extremely efficient
% to process---for example, computing Tajima's $D$ using \tskit\
% is several orders of magnitude faster for large sample sizes than
% efficient libraries working with the variant matrix
% representation~\citep{ralph2020efficiently}.
% \jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
% here. Same basic idea, but much cruder.}

@@

\jkcomment{A point to make somewhere in here: the tskit data model is
integrated, with all of the information that you need in one file. Not
splattered across a number of different files, that you have to keep track
of.}

@@

\subsection*{Simulating mutations}
\label{sec-mutations}
% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees.
The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format. Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
The output of the combined simulations of ancestry and mutations is
then some representation of the resulting nucleotide sequences
(or perhaps some derived statistics of those sequences);
information about the mutations themselves is not typically
provided to the user.

The approach taken in \msprime\ and \tskit\ is quite different.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
integral part of the tree sequence data model, and genetic
variation is encoded by sites and the mutations at those sites.
This may seem at first inconvenient, but is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we required $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space~\citep{kelleher2016efficient}. For
large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster for large
sample sizes~\citep{ralph2020efficiently}.

Motivated by the large numbers of SARS-CoV-2 genomes currently
available, \cite{turakhia2021ultrafast} propose a similar approach
to encoding variation data for non-recombining organisms,
where a Newick tree is annotated by mutations~\citep{mcbroome2021daily}.
Similarly, the Phastsim~\citep{demaio2021phastsim} mutation simulator,
developed primarily to support simulation of SARS-CoV-2 trees
with hundreds of thousands of leaves, can optionally output a
Newick tree, where nodes are annotated with the mutations.
However, the dependence on the Newick format for encoding
topological information means that it suffers from all the
problems of newick [what?], and there is no software library support
[confirm].

\jkcomment{stuff to cover somewhere in this section}
\begin{itemize}
\item Mutation simulations is only included in msprime for historical reasons.
It would be better to see msprime's mutation simulation as a separate
package. The mutation and ancestry code in msprime have no priveledged
access to each other, and mutation simulation would be no less efficient
if done in a different package.
\item Really emphasise this point: you can use msprime's mutation simulation
on \emph{any} tskit tree sequence, produced by SLiM, fwdpp, etc or
even tsinfer. It's completely modular.
\end{itemize}

Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

\subsection*{Selection}
\label{sec-selection}
Another ellaboration of the standard neutral coalescent with recombination
is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
In the context of the coalescent this amounts to creating a structured population
during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
in which lineages may transit between favored and unfavored backgrounds through
recombination. This allows for many selective sweep scenarios, including recurrent, partial,
and soft selective sweeps to be simulated efficiently, however in general this
efficiency comes at the cost of flexibility in comparison to forwards in
time simulation.
